Avoidance in permutations

### 2014-11-07 project2/paper/paper.tex

With examples of a four element sequence that avoids 123 and a four element sequence that does not avoid 123, it is natural to ask the question: how many four element sequences avoid 123? And more generally, how many $n$-element sequences avoid 123? If we define $s_n(123)$ to be the number of $n$ element sequences that avoid 123, then the question above is equivalent to identifying the sequence $(s_n(123))_n$. In this paper we consider specific cases of a very general avoidance problem, describing the sequence $(s_n(\sigma))_n$ where $\sigma$ is any permutation of the elements of the set $\{1,2,\cdots, k\}\subset\mathbb{N}$ for some $k$.

In Section \ref{S3} we consider sequences $s_n(\pi)$ when $\pi$ is a permutation on the set $\{1,2,3\}$, and prove that the sequence $s_n(\pi)$ for all such $\pi$ is in fact the sequence of Catalan numbers. In Section \ref{S4} we will consider sequences $s_n(\pi)$ when $\pi$ is a permutation of $\{1,2,3,4\}$. In Section \ref{Tn} we consider a variation of the avoidance problem that considers the problem of avoiding sequences in a subset $T_n$ of the permutations of $n$ elements. Before proceeding, however, we formally introduce the notation that will be used through the rest of this paper.

@@

\begin{remark}
The notation used to describe the elements of $S_n$ above is `one-line' notation.\todo{Jerison wanted us to define this formally, I think} Thus far we have ignored the fact that $S_n$ has a group structure. When discussing this group structure it is convenient to use a different notation known as `cycle' notation.\todo{Also should define this formally} This alternative notation will be used in Section \ref{S4} and an introduction to cycle notation can be found in Appendix [].

The notation used to describe the elements of $S_n$ above is `one-line' notation. Thus far we have ignored the fact that $S_n$ has a group structure. When discussing this group structure it is convenient to use a different notation known as `cycle' notation. This alternative notation will be used in Section \ref{S4} and an introduction to cycle notation can be found in Appendix []. 
\end{remark}

\begin{definition}
Two finite sequences of distinct positive integers $a_1,\cdots a_k$ and $b_1,\cdots b_k$ have the same relative order if the $i^\text{th}$ largest entry of each unordered set $\{a_1,\cdots,a_k\}$ and $\{b_1,\cdots, b_k\}$ appears in the same position in $a_1\cdots a_k$ and $b_1\cdots b_k$ for each $1\le i\le k$. 
\end{definition}\todo{I changed this definition... take a look and see if you agree with it}

>>>>

With examples of a four element sequence that avoids 123 and a four element sequence that does not avoid 123, it is natural to ask the question: how many four element sequences avoid 123? And more generally, how many $n$-element sequences avoid 123? For a sequence $\pi$ that is a permutation of the elements of $\{1,\cdots,k\}$, we define
$$
s_n(\pi)=\#\{\text{the set of permutations of $\{1,\cdots, n\}$ that avoid $\pi$}\}.
$$

If we define $s_n(123)$ to be the number of $n$ element sequences that avoid 123, then the question above is equivalent to identifying the sequence $(s_n(123))_n$. In this paper we consider specific cases of a very general avoidance problem, describing the sequence $(s_n(\sigma))_n$ where $\sigma$ is any permutation of the elements of the set $\{1,2,\cdots, k\}\subset\mathbb{N}$ for some $k$.

In Sections \ref{S3} and \ref{S4} we consider sequences $s_n(\pi)$ for $\pi$ a permutation on the set $\{1,2,3\}$, and prove that the sequence $s_n(\pi)$ for all such $\pi$ is in fact the sequence of Catalan numbers. In Section \ref{S4} we will consider sequences $s_n(\pi)$ for $\pi$ a permutation of $\{1,2,3,4\}$. In Section \ref{Tn} we consider a variation of the avoidance problem that considers the problem of avoiding sequences in a subset $T_n$ of the permutations of $n$ elements. Before proceeding, however, we formally introduce the notation that will be used throughout this paper.

@@

%Let $S_n$ be the set of permutations of $\{1,\cdots, n\}$. There are two conventions commonly used to write elements of $S_n$. The first is $\emph{one-line}$ notation where permutations in $S_n$ are written explicitly, e.g. 43215 is a permutation of $\{1,2,3,4,5\}$ in $S_5$.

%The second is \emph{cycle notation}, where a cycle $(a_1\cdots a_n)\in S_n$ with $1\le a_i\le n$ acts on a sequence by sending $a_1\mapsto a_2$, $a_2\mapsto a_3$, $\cdots$, $a_n\mapsto a_1$. For example, the cycle $(14)(23)$ sends $12345$ to $43215$.

\begin{definition}
Two finite sequences of distinct positive integers $a_1,\cdots a_k$ and $b_1,\cdots b_k$ have the same relative order if the $i^\text{th}$ largest entry of each unordered set $\{a_1,\cdots,a_k\}$ and $\{b_1,\cdots, b_k\}$ appears in the same position in $a_1\cdots a_k$ and $b_1\cdots b_k$ for each $1\le i\le k$. 
\end{definition}
